Cooling Requirements for the Vertical Shear Instability in Protoplanetary Disks

### 2015-04-24 appendix.tex
minor notation corrections

% In \S\ref{approx_gov} we made the replacement $D\to\Omega_K^2$ before
% eliminating variables to obtain a single equation for $\delta v_z$ in
% vertically isothermal disks, Eq. \ref{vertiso_gov}.  
% This procedure ignores the vertical dependence of
% $D=\kappa^2(z) - \sigma^2$. We show here that this has no serious 
% consequence for thin disks.

@@

\begin{align}\label{full_ode}
  0 =& \frac{d^2\delta v_z}{dz^2} + \left[1 + \frac{\ii k_x c_s^2
      q}{Dr} - \frac{k_x^2c_s^2}{\left(k_x^2c_s^2 + \chi
        D\right)}\frac{qc_s^2F}{Dr^2}\right]\frac{d\ln\rho}{dz}\frac{d\delta
    v_z}{dz} \notag\\
  &+ \left\{\upsilon^2\left(\frac{k_x^2}{D} +
      \frac{\chi}{c_s^2}\right) + \left(\chi + \frac{\ii k_x c_s^2
        q}{Dr}\right)\frac{d^2\ln\rho}{dz^2} -
    \frac{c_s^2}{D}\left(\frac{d\ln\rho}{dz}\right)^2\left(k_x^2 -
      \frac{\ii k_x q}{r}\right)
   \left[\left(1-\chi\right) +
     \frac{\chi}{\left(k_x^2c_s^2 + \chi D\right)}\frac{qc_s^2 F}{r^2}\right] 
   \right\}\delta v_z,
\end{align}
where we recall $\chi = \left(1-\ii\upsilon t_c\right)/\left(1-\ii\upsilon
t_c \gamma\right)$. Making the low frequency and Keplerian
approximations is equivalent to the replacement $D\to
\Omega_K^2$. This gives, in terms of dimensionless variables,  
\begin{align}
  &\delta v_z ^{\prime\prime} + \left[1 + \ii h q\hat{k} -
    \frac{ \hat{k}^2}{
      \left(\hat{k}^2+\chi\right)}q h^2F\right]\ln\rho^{\prime}\delta v_z^\prime +
  \left\{\left(\chi + \ii h q
      \hat{k}\right)\ln\rho^{\prime\prime} - \ln\rho^{\prime
      2}\left(\khat^2 -
      \ii h
      q\hat{k}\right)\left[1 - \chi +
      \frac{\chi}{\left(\hat{k}^2+\chi\right)}qh^2F\right]\right\}\delta v_z \notag\\&=
  -\hat{\upsilon}^2\left(\hat{k}^2+\chi\right)\delta v_z\label{adia_diso3}.
\end{align} 
Since we are interested in thin disks with $h \ll 1$ and $|q| =
O(1)$, we neglect the terms proportional to $h^2$, giving 
\begin{align}
  0= \delta v_z ^{\prime\prime} + \left(1 + \ii h
     q\hat{k}\right)\ln\rho^{\prime}\delta v_z^\prime 
   +
   \left[\hat{\upsilon}^2\left(\hat{k}^2+\chi\right) +
     \left(  \chi + \ii h q\hat{k}\right)\ln\rho^{\prime\prime}
     - \ln\rho^{\prime
       2}\left(\khat^2 -
       \ii h
       q\hat{k}\right)\left(1 - \chi\right)\right]\delta v_z,\label{vertiso_gov_nondim}
 \end{align}
 which is used in the discussion below.  Eq. \ref{nearly_iso_explicit}, as
 used in our analytical discussion, is obtained by inserting the 
 thin-disk expression for the density field. We comment that the simpler way
 of obtaining Eq. \ref{vertiso_gov_nondim} is to neglect the vertical 
 dependence of $D$ by setting $D\to\Omega_K^2$ in Eq. \ref{ode_w} before eliminating
 variables in favor of $\delta v_z$.

@@

%Eq. \ref{adia_diso3} differs from Eq. \ref{vertiso_gov_nondim} by terms
%proportional to $h^2$. For a thin disk ($h\ll1$) with
%$|q|=O(1)$, as considered in this work, neglecting these terms has no
%qualitative effect on the discussion in \S\ref{analytic_relax}. 
% In deriving Eq. \ref{vertiso_gov}, the vertical dependence of
% $D=\kappa^2(z)-\sigma^2$ was ignored because the replacement
% $D\to\Omega_K^2$ was made before eliminating variables in favor of
% $\delta v_z$. In Appendix \ref{adia_improve} we show that making this
% replacement after eliminating variables, which involves $dD/dz$,
% introduces terms of $O(h^2)$. Since we are interested in 
% thin disks ($h\ll 1 $), e.g. protoplanetary disks where 
% $h\lesssim 0.1$ \citep{chiang10}, these terms introduce
% unnecessary complexity for the present discussion.

@@

\section{Linear problem in the isothermal limit}\label{iso_discuss}  
Here we summarize selected results for isothermal
perturbations ($\beta\equiv 0$) in vertically isothermal disks 
($\Gamma=1$) in the VGSB framework. In this case Eq. \ref{ode_Q}
becomes    
\begin{align} 
  Q = W 
\end{align}
(i.e. $\delta P = c_s^2\delta \rho$)% , and so $  \ii\hat{\upsilon}c_s\delta v_z = W^\prime $
% from Eq. \ref{lin_vz}
. For isothermal perturbations it is 
simpler to work with an equation for $W$ by substituting $Q=W$ into
Eq. \ref{ode_w} and using Eq. \ref{ode_vz} to eliminate $\delta v_z$. 
In this case, we shall not yet make the low frequency
approximation, but first make the  Keplerian approximation. We
obtain, in terms of dimensionless variables, 
\begin{align}
  0 = W^{\prime\prime} + \left[\ln\rho^\prime - \frac{\ii h q \hat{k} f(\zhat)
      }{1-\hat{\upsilon}^2} \right]W^\prime +
  \hat{\upsilon}^2\left(1+\frac{\hat{k}^2}{1-\hat{\upsilon}^2}\right)W,\label{iso_ode0} 
\end{align}
where for discussion purposes we have defined $f(\zhat)$ such that
\begin{align}\label{fz_shear}
  \frac{d\Omega^2}{d\hat{z}} = h^2q f(\hat{z})\Omega_K^2.
\end{align}
By comparing Eq. \ref{fz_shear} with Eq. \ref{vertical_shear_ex}, we
see that $f(\zhat)= 
\zhat\left(1+h^2\zhat^2\right)^{-3/2}$. More generally, though,
$f$ may be regarded as a representation of the vertical
shear profile. Physically, we expect there is a maximum value of 
$|d\Omega/dz|$, the existence of which should limit the growth rate of
the VSI,  as remarked in \S\ref{vshear_def}. We explicitly demonstrate
this below.

@@

In practice, one might consider a vertical domain of a few scale 
heights in a thin disk with $|q|=O(1)$. In this case, $|h\zhat|\ll1$ and  
$f\simeq \zhat$, so that $\mathrm{max}|f| = O(1)$, implying a
maximum growth rate $O(h \Omega_K)$, consistent with numerical
results.

@@

% \subsubsection{Stability in the absence of vertical
%   shear}\label{iso_stable}
% \
% Eq. \ref{iso_ode3} is
% \begin{align}\label{hermite_ode}
%   \left[w(\hat{z})W^\prime \right]^\prime + nW
%   w(\hat{z}) =0, 
% \end{align}
% where $w(\zhat)=\exp{(-\zhat^2/2)}$ and we have defined a new eigenvalue 
% \begin{align}
%   n \equiv \hat{\upsilon}^2(1+\hat{k}^2). 
% \end{align} 
% Eq. \ref{hermite_ode} is Hermite's differential equation. If we impose
% that the kinetic energy density remain bounded at infinity, then the
% solutions are Hermite polynomials  
% \begin{align}
%   W \propto \He_n(\hat{z}),
% \end{align}
% and $n$ is a non-negative integer. This translates to a real
% eigenfrequency $\sigma$ such that
% \begin{align}
%   \left|\hat{\upsilon}\right| = \sqrt{n}
%   \left(1+\hat{k}^2\right)^{-1/2}. \label{iso_stable_disp}
% \end{align}
% Since we have assumed $|\sigma^2|\ll \kappa^2\sim \Omega_K^2$, our
% analysis is only valid for large wavenumbers such that $\hat{k}^2\gg   
% n-1$. Physically, this corresponds to radial length-scales much
% smaller than the local disk scale height. Note that
% Eq. \ref{iso_stable_disp} is implied by Eq. 54 in
% \cite{lubow93} after setting $\gamma=1$ and making the low-frequency,
% nearly-Keplerian approximation in that equation.

>>>>

% In \S\ref{approx_gov} we made the replacement $D\to\OmK^2$ before
% eliminating variables to obtain a single equation for $\delta v_z$ in
% vertically isothermal disks, Eq. \ref{vertiso_gov}.  
% This procedure ignores the vertical dependence of
% $D=\kappa^2(z) - \sigma^2$. We show here that this has no serious 
% consequence for thin disks.

@@

\begin{align}\label{full_ode}
  0 =& \frac{d^2\delta v_z}{dz^2} + \left[1 + \frac{\ii k_x c_s^2
      q}{Dr} - \frac{k_x^2c_s^2}{\left(k_x^2c_s^2 + \chi
        D\right)}\frac{qc_s^2F}{Dr^2}\right]\frac{d\ln\rho}{dz}\frac{d\delta
    v_z}{dz} \notag\\
  &+ \left\{\upsilon^2\left(\frac{k_x^2}{D} +
      \frac{\chi}{c_s^2}\right) + \left(\chi + \frac{\ii k_x c_s^2
        q}{Dr}\right)\frac{d^2\ln\rho}{dz^2} -
    \frac{c_s^2}{D}\left(\frac{d\ln\rho}{dz}\right)^2\left(k_x^2 -
      \frac{\ii k_x q}{r}\right)
   \left[\left(1-\chi\right) +
     \frac{\chi}{\left(k_x^2c_s^2 + \chi D\right)}\frac{qc_s^2 F}{r^2}\right] 
   \right\}\delta v_z,
\end{align}
where we recall $\chi = \left(1-\ii\upsilon t_c\right)/\left(1-\ii\upsilon
t_c \gamma\right)$. Making the low frequency and Keplerian
approximations is equivalent to the replacement $D\to
\OmK^2$. This gives, in terms of dimensionless variables,  
\begin{align}
  &\delta v_z ^{\prime\prime} + \left[1 + \ii h q\hat{k} -
    \frac{ \hat{k}^2}{
      \left(\hat{k}^2+\chi\right)}q h^2F\right]\ln\rho^{\prime}\delta v_z^\prime +
  \left\{\left(\chi + \ii h q
      \hat{k}\right)\ln\rho^{\prime\prime} - \ln\rho^{\prime
      2}\left(\khat^2 -
      \ii h
      q\hat{k}\right)\left[1 - \chi +
      \frac{\chi}{\left(\hat{k}^2+\chi\right)}qh^2F\right]\right\}\delta v_z \notag\\&=
  -\hat{\upsilon}^2\left(\hat{k}^2+\chi\right)\delta v_z\label{adia_diso3}.
\end{align} 
Since we are interested in thin disks with $h \ll 1$ and $|q| =
O(1)$, we neglect the terms proportional to $h^2$, giving 
\begin{align}
  0= \delta v_z ^{\prime\prime} + \left(1 + \ii h
     q\hat{k}\right)\ln\rho^{\prime}\delta v_z^\prime 
   +
   \left[\hat{\upsilon}^2\left(\hat{k}^2+\chi\right) +
     \left(  \chi + \ii h q\hat{k}\right)\ln\rho^{\prime\prime}
     - \ln\rho^{\prime
       2}\left(\khat^2 -
       \ii h
       q\hat{k}\right)\left(1 - \chi\right)\right]\delta v_z,\label{vertiso_gov_nondim}
 \end{align}
 which is used in the discussion below.  Eq. \ref{nearly_iso_explicit}, as
 used in our analytical discussion, is obtained by inserting the 
 thin-disk expression for the density field. We comment that the simpler way
 of obtaining Eq. \ref{vertiso_gov_nondim} is to neglect the vertical 
 dependence of $D$ by setting $D\to\OmK^2$ in Eq. \ref{ode_w} before eliminating
 variables in favor of $\delta v_z$.

@@

%Eq. \ref{adia_diso3} differs from Eq. \ref{vertiso_gov_nondim} by terms
%proportional to $h^2$. For a thin disk ($h\ll1$) with
%$|q|=O(1)$, as considered in this work, neglecting these terms has no
%qualitative effect on the discussion in \S\ref{analytic_relax}. 
% In deriving Eq. \ref{vertiso_gov}, the vertical dependence of
% $D=\kappa^2(z)-\sigma^2$ was ignored because the replacement
% $D\to\OmK^2$ was made before eliminating variables in favor of
% $\delta v_z$. In Appendix \ref{adia_improve} we show that making this
% replacement after eliminating variables, which involves $dD/dz$,
% introduces terms of $O(h^2)$. Since we are interested in 
% thin disks ($h\ll 1 $), e.g. protoplanetary disks where 
% $h\lesssim 0.1$ \citep{chiang10}, these terms introduce
% unnecessary complexity for the present discussion.

@@

\section{Linear problem in the isothermal limit}\label{iso_discuss}  
Here we summarize selected results for isothermal
perturbations ($\beta\equiv 0$) in vertically isothermal disks 
($\Gamma=1$) in the VGSB framework. In this case Eq. \ref{ode_Q}
becomes    
\begin{align} 
  Q = W 
\end{align}
(i.e. $\delta P = c_s^2\delta \rho$)% , and so $  \ii\hat{\upsilon}c_s\delta v_z = W^\prime $
% from Eq. \ref{lin_vz}
. For isothermal perturbations it is 
simpler to work with an equation for $W$ by substituting $Q=W$ into
Eq. \ref{ode_w} and using Eq. \ref{ode_vz} to eliminate $\delta v_z$. 
In this case, we shall not yet make the low frequency
approximation, but first make the  Keplerian approximation. We
obtain, in terms of dimensionless variables, 
\begin{align}
  0 = W^{\prime\prime} + \left[\ln\rho^\prime - \frac{\ii h q \hat{k} f(\zhat)
      }{1-\hat{\upsilon}^2} \right]W^\prime +
  \hat{\upsilon}^2\left(1+\frac{\hat{k}^2}{1-\hat{\upsilon}^2}\right)W,\label{iso_ode0} 
\end{align}
where for discussion purposes we have defined $f(\zhat)$ such that
\begin{align}\label{fz_shear}
  \frac{d\Omega^2}{d\hat{z}} = h^2q f(\hat{z})\OmK^2.
\end{align}
By comparing Eq. \ref{fz_shear} with Eq. \ref{vertical_shear_ex}, we
see that $f(\zhat)= 
\zhat\left(1+h^2\zhat^2\right)^{-3/2}$. More generally, though,
$f$ may be regarded as a representation of the vertical
shear profile. Physically, we expect there is a maximum value of 
$|d\Omega/dz|$, the existence of which should limit the growth rate of
the VSI,  as remarked in \S\ref{vshear_def}. We explicitly demonstrate
this below.

@@

In practice, one might consider a vertical domain of a few scale 
heights in a thin disk with $|q|=O(1)$. In this case, $|h\zhat|\ll1$ and  
$f\simeq \zhat$, so that $\mathrm{max}|f| = O(1)$, implying a
maximum growth rate $O(h \OmK)$, consistent with numerical
results.

@@

% \subsubsection{Stability in the absence of vertical
%   shear}\label{iso_stable}
% \
% Eq. \ref{iso_ode3} is
% \begin{align}\label{hermite_ode}
%   \left[w(\hat{z})W^\prime \right]^\prime + nW
%   w(\hat{z}) =0, 
% \end{align}
% where $w(\zhat)=\exp{(-\zhat^2/2)}$ and we have defined a new eigenvalue 
% \begin{align}
%   n \equiv \hat{\upsilon}^2(1+\hat{k}^2). 
% \end{align} 
% Eq. \ref{hermite_ode} is Hermite's differential equation. If we impose
% that the kinetic energy density remain bounded at infinity, then the
% solutions are Hermite polynomials  
% \begin{align}
%   W \propto \He_n(\hat{z}),
% \end{align}
% and $n$ is a non-negative integer. This translates to a real
% eigenfrequency $\sigma$ such that
% \begin{align}
%   \left|\hat{\upsilon}\right| = \sqrt{n}
%   \left(1+\hat{k}^2\right)^{-1/2}. \label{iso_stable_disp}
% \end{align}
% Since we have assumed $|\sigma^2|\ll \kappa^2\sim \OmK^2$, our
% analysis is only valid for large wavenumbers such that $\hat{k}^2\gg   
% n-1$. Physically, this corresponds to radial length-scales much
% smaller than the local disk scale height. Note that
% Eq. \ref{iso_stable_disp} is implied by Eq. 54 in
% \cite{lubow93} after setting $\gamma=1$ and making the low-frequency,
% nearly-Keplerian approximation in that equation.

### 2015-04-24 setup.tex
minor notation corrections

 Perturbations take the form of a radial plane wave with arbitrary vertical dependence, 
 e.g.\
 \begin{align}
  \delta\rho \rightarrow \delta\rho(z)\exp{\left(\ii k_x x - \ii\upsilon
      t\right)},    
\end{align}
where $\zeta = \omega + \ii \sigma$ is a (generally) complex frequency, 
with $\omega$ and $\sigma$ being
the real frequency and growth rate, respectively, and $k_x$ is a real 
radial wavenumber.  Henceforth all unperturbed fluid variables, including 
gradients such as $\p \Omega/ \p z$, refer to  equilibrium values at $(r_0, z)$.

We further define the perturbation variables  $W \equiv \delta P /\rho$ 
and $Q \equiv c_s^2\delta\rho/\rho$.  With this procedure the linearized
system of equations are
\begin{align}
  \ii\upsilon \frac{Q}{c_s^2}  &=  \ii k_x \delta v_x + \frac{d\dd
    v_z}{dz} + \frac{\p\ln{\rho}}{\p z}\delta v_z + \zeta
  \frac{\p\ln{\rho}}{\p r} \delta v_x,\label{lin_mass}\\
  \ii\upsilon \delta v_x  &= \ii k_x W - 2\Omega\delta v_y -
  \zeta\frac{1}{\rho}\frac{\p P}{\p r}\frac{Q}{c_s^2}\label{lin_vx}\\
   \ii \upsilon\delta v_y &= r\frac{\p \Omega}{\p z}\delta v_z +
  \frac{\kappa^2}{2\Omega}\delta v_x, \label{lin_vy}\\
   \ii\upsilon\delta v_z &= \frac{dW}{dz} +
  \frac{\p\ln{\rho}}{\p z}\left(W-Q\right), \label{lin_vz}\\
  \ii \upsilon W &= c_s^2\frac{\p\ln{\rho}}{\p z}\delta v_z +
  c_s^2\frac{\gamma}{\Gamma} \left(\ii k_x\delta v_x + \frac{d\delta
      v_z}{dz}\right) \notag\\
  &+ \frac{1}{t_c}\left(W - \frac{Q}{\Gamma}\right) +
  \zeta\frac{1}{\rho}\frac{\p P}{\p r}\delta v_x.\label{lin_energy}
  % -&\ii\upsilon W + \frac{1}{t_c}\left(W-\frac{Q}{\Gamma}\right) = -\ii\upsilon\frac{\gamma}{\Gamma} Q +
  % c_s^2\frac{d\ln{\rho}}{dz}\left(\frac{\gamma}{\Gamma}-1\right)\delta v_z.\label{lin_energy}
\end{align}
The above equations are a set of ODEs (ordinary differential equations) in $z$.  Solutions to these 
are presented in \S\ref{numerical} and \S\ref{application}.   The coefficient $ \zeta = 1$ is 
introduced simply to label terms containing radial gradients of the equilibrium state, 
which again are evaluated at $(r_o, z)$.   We will consider the effects of ignoring these radial gradients below.

>>>>

 Perturbations take the form of a radial plane wave with arbitrary vertical dependence, 
 e.g.\
 \begin{align}
  \delta\rho \rightarrow \delta\rho(z)\exp{\left(\ii k_x x - \ii\upsilon
      t\right)},    
\end{align}
where $\upsilon = \omega + \ii \sigma$ is a (generally) complex frequency, 
with $\omega$ and $\sigma$ being
the real frequency and growth rate, respectively, and $k_x$ is a real 
radial wavenumber.  Henceforth all unperturbed fluid variables, including 
gradients such as $\p \Omega/ \p z$, refer to  equilibrium values at $(r_0, z)$.

We further define the perturbation variables  $W \equiv \delta P /\rho$ 
and $Q \equiv c_s^2\delta\rho/\rho$.  With this procedure the linearized
system of equations are
\begin{align}
  \ii\upsilon \frac{Q}{c_s^2}  &=  \ii k_x \delta v_x + \frac{d\dd
    v_z}{dz} + \frac{\p\ln{\rho}}{\p z}\delta v_z + \zeta
  \frac{\p\ln{\rho}}{\p r} \delta v_x,\label{lin_mass}\\
  \ii\upsilon \delta v_x  &= \ii k_x W - 2\Omega\delta v_y -
  \zeta\frac{1}{\rho}\frac{\p P}{\p r}\frac{Q}{c_s^2}\label{lin_vx}\\
   \ii \upsilon\delta v_y &= r\frac{\p \Omega}{\p z}\delta v_z +
  \frac{\kappa^2}{2\Omega}\delta v_x, \label{lin_vy}\\
   \ii\upsilon\delta v_z &= \frac{dW}{dz} +
  \frac{\p\ln{\rho}}{\p z}\left(W-Q\right), \label{lin_vz}\\
  \ii \upsilon W &= c_s^2\frac{\p\ln{\rho}}{\p z}\delta v_z +
  c_s^2\frac{\gamma}{\Gamma} \left(\ii k_x\delta v_x + \frac{d\delta
      v_z}{dz}\right) \notag\\
  &+ \frac{1}{t_c}\left(W - \frac{Q}{\Gamma}\right) +
  \zeta\frac{1}{\rho}\frac{\p P}{\p r}\delta v_x.\label{lin_energy}
  % -&\ii\upsilon W + \frac{1}{t_c}\left(W-\frac{Q}{\Gamma}\right) = -\ii\upsilon\frac{\gamma}{\Gamma} Q +
  % c_s^2\frac{d\ln{\rho}}{dz}\left(\frac{\gamma}{\Gamma}-1\right)\delta v_z.\label{lin_energy}
\end{align}
The above equations are a set of ODEs (ordinary differential equations) in $z$.  Solutions to these 
are presented in \S\ref{numerical} and \S\ref{application}.   The coefficient $ \zeta = 1$ is 
introduced simply to label terms containing radial gradients of the equilibrium state, 
which again are evaluated at $(r_0, z)$.   We will consider the effects of ignoring these radial gradients below.
